Exploratory Data Analysis on White Wine by Chesaney Wyse
========================================================


```{r message=FALSE, warning=FALSE, include=FALSE}

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.
library(ggplot2)
library(scales)
library(memisc)
library(GGally)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data

wine <- read.csv('wineQualityWhites.csv')

```

This report aims to identify the variables that are most likely to predict the 
quality of white wine. This predicion will be established utilizing a dataset 
which consists of the quality ratings of approximately 5000 white wines. The 
dataset contains 11 variables measuring several chemical properties of the wine.
This report will also explore the relationships between the chemical properties
to determine which properties affect other properties. 

# Univariate Plots Section

```{r echo=FALSE}
str(wine)
```

```{r echo=FALSE}
summary(wine)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x=fixed.acidity, data=wine)
```

There appear to be a few outliers for fixed acidity on the higher end of the 
spectrum. The maximum value for fixed acidity is 14.2, which appears to be very 
high, as the typical range is around 5-8. 

```{r echo=FALSE, message=FALSE}
qplot(x=volatile.acidity, data=wine)
```

Volatile acidity has a similar-looking distribuution as fixed acidity. The 
typical range is from 0.25-0.35. This variable also has some higher-value 
outliers, giving the data a right skew. I am curious to see what quality rating 
these outliers recieved. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x=citric.acid, data=wine)
```

The bulk of white wines have a citric acid content of approximately 0.2-0.4 
g/dm^3. The maximum value for citric acid is 1.66 g/dm^3. I know that citric 
acid can create a sour taste, so I am curious how this would affect the quality 
rating, and if more or less acidic wines are prefered by wine experts. While the 
outliers are easy to identify with the measurements provided, I wonder how 
prevalent the effect of the citric acid is to the wine experts. 

```{r echo=FALSE, message=FALSE, warning=TRUE}
qplot(x=residual.sugar, data=wine)
```

A lower amount of residual sugar is clearly more common for white wines. 
However, there are a few outliers with much greater amounts of residual sugar. 
As this is clearly not typical of white wines, I am curious how a higher amount 
of sugar would affect the quality rating. 

```{r echo=FALSE, message=FALSE}
qplot(x=chlorides, data=wine)
```
```{r echo=FALSE, message=FALSE}
qplot(x=chlorides, data=wine)+
  scale_x_log10()
```

Chlorides is a right-skewed variable, with the bulk of the measurements falling 
between 0.03 and 0.06. The second histogram shows the variable on a log scale. 

```{r echo=FALSE, message=FALSE}
qplot(x=free.sulfur.dioxide, data=wine)
```
```{r echo=FALSE, message=FALSE}
qplot(x=free.sulfur.dioxide, data=wine)+
  scale_x_log10()
```

Free sulfur dioxide is another variable with a significant right skew, so I 
log-transformed the data to better view the distribution. On a log scale, the 
distribution appears normal. 

```{r echo=FALSE, message=FALSE}
qplot(x=total.sulfur.dioxide, data=wine)
```

Total sulfur dioxide has a similar distribtuion to free sulfur dioxide. I am 
assuming these two variables will have a pretty strong correlation. As the free 
sulfur dioxide increases, it would make sense that the total sulfur dioxide 
measure would increase as well. 

```{r echo=FALSE, message=FALSE}
qplot(x=density, data=wine)
```
```{r}
summary(wine$density)
```

The majority of white wines in this sample have a density between .99 and 1.0. 
There is a slight right skew to the variable, as there are some wines with 
densities above 1.0. These are clearly outside of the typical range as displayed 
in the above histogram. 

```{r echo=FALSE, message=FALSE}
qplot(x=pH, data=wine)
```

```{r echo=FALSE}
summary(wine$pH)
```

The variable pH appears to be normally distributed. As indicated by the range of 
values, white wine is an acidic substance, with an average pH of 3.188. It will 
be interesting to see which pH has the highest quality rating, as well as the 
relationship between pH and citric acid content. 

```{r echo=FALSE, message=FALSE}
qplot(x=sulphates, data=wine)
```

The variable sulphates has a bimodal appearance, with a tail off to the right of 
the graph illustrating the right skew in the data. 

```{r echo=FALSE, message=FALSE}
qplot(x=alcohol, data=wine)
```

There is a lot of variance in the histogram for alcohol content, as the alcohol 
content is measured to one decimal place. However, it does appear that most 
white wines fall between 9 and 11 percent.  

```{r echo=FALSE, message=FALSE}
qplot(x=quality, data=wine)
```

```{r}
summary(wine$quality)
```

The most common quality rating for the sample is 6. The quality ratings appear 
to be pretty evenly distributed, with no significant skew. 



# Univariate Analysis

### What is the structure of your dataset?

There are 4,898 white wines in the dataset, with 11 variables measuring 
different chemical properties of the wine (fixed acidity, volatile acidity, 
citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur 
dioxide, density, pH, sulfates, and alcohol). The output variable, quality, is 
based on a scale of 0 to 10, zero being the worst and ten being the best. These 
quality ratings were determined by at least three wine experts. 

### What is/are the main feature(s) of interest in your dataset?

Upon researching wine tasting, I found that the main variables that would likely 
influence the quality rating are pH, alcohol content, and residual sugar. I will 
be exploring the effects of these variables on the quality rating. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I would also like to explore the effect density has on the quality rating, as 
the density of wine is reflective of the alcohol and residual sugar content. 
Additionally, I would like to explore how well the acidic measurements (fixed 
acidity, volatile acidity, and citric acid)  explain the pH of the wine. 

### Did you create any new variables from existing variables in the dataset?

Yes, I cleaned up the alcohol variable by cutting it into the new variable: 
alcohol_bucket. This eliminated some of the noise in the alcohol histogram by 
organizing the data into whole number percentages, rather than to a decimal 
point. The purpose of this transformation was to identify the most typical 
alcohol percentage for white wines. 

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free 
sulfur dioxide, total sulfur dioxide, and density all are right-skewed. I 
examined a few of these on a log scale to limit the impact of the skew on the 
visual. 

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=TRUE}
wine_subset <- wine[,c(2:13)]
ggpairs(wine_subset)
```
```{r echo=FALSE}
cor(wine_subset)
```

There are not many strong relationships in the dataset. The strongest 
correlation is between density and residual sugar, with a pearson correlation 
coefficient of 0.83896645. Of all the variables, alcohol has the strongest 
correlation coefficient with quality. 

```{r echo=FALSE}
ggplot(data=wine, aes(x=alcohol, y=quality))+
  geom_point(alpha=0.05)
```

In this particular scatterplot, it is difficult to identify whether or not there 
is a relationship between alcohol content and quality rating.  To better examine 
this realtionship, we can look at the average quality rating by alcohol content. 


```{r echo=FALSE}
ggplot(data=wine, aes(x=alcohol, y=quality))+
  geom_line(stat='summary', fun.y=mean)
```

Here, we can identify a general upward trend. It appears that as the alcohol 
content increases, the average quality rating increases as well. However, there 
seems to be a lot of variation amongst the middle-range alcohol content wines. 
To explore this issue, the alcohol variable can be cut so that the values are 
grouped within single increment integers. 

```{r echo=FALSE}
wine$alcohol_bucket <- cut(wine$alcohol, c(8,9,10,11,12,13,14,15))
table(wine$alcohol_bucket)
```
```{r}
summary(wine$alcohol)
```

It does appear that there is are a greater number of wines from the dataset that 
fall within this region of high variation. To eliminate some of the noise, the 
realtionship between alcohol and quality can be viewed in a graph using the cut 
variable version of alcohol content.  

```{r echo=FALSE}
ggplot(data=subset(wine, !is.na(alcohol_bucket)), 
       aes(x=alcohol_bucket, y=quality))+
  geom_point(stat='summary', fun.y=mean)
```

In this plot, it is much easier to observe the upward trend in the data. In 
fact, there is only one region where the quality decreases as the alcohol 
content increases (between the first and second buckets). 

```{r echo=FALSE}
ggplot(data=wine, aes(x=citric.acid, y=quality))+
  geom_point(alpha=.1)
```
```{r echo=FALSE}
ggplot(data=wine, aes(x=citric.acid, y=quality))+
  geom_point(stat='summary', fun.y=mean)
```

During some research on wine quality, I discovered that acidity is a key 
characteristic when tasting wine. I decided to take a look at the relationship 
between citric acid content and quality rating. I was unable to discover a 
significant realtionship, although there are two very low points in the plot, 
which I find intersting. Perhaps pH provides a stronger relationship. 

```{r echo=FALSE}
ggplot(data=wine, aes(x=pH, y=quality))+
  geom_point()
```
```{r echo=FALSE}
ggplot(data=wine, aes(x=pH, y=quality))+
  geom_line(stat='summary', fun.y=mean)
```
```{r}
summary(wine$pH)
```

The variable pH extends to three decimal places, which may explain some of the 
noise present in the graph. It may be beneficial to cut this variable as done 
with the alcohol variable. 

```{r}
wine$ph_buckets <- cut(wine$pH, c(2.7, 2.9, 3.1, 3.3, 3.5, 3.7, 3.9))
summary(wine$ph_buckets)
```

```{r echo=FALSE}
ggplot(data=subset(wine, !is.na(ph_buckets)), 
       aes(x=ph_buckets, y=quality))+
  geom_point(stat='summary', fun.y=mean)
```

Although there is not a clear linear relationship between the pH and quality of 
the wine, it is evident that on average, white wines with a pH within the range 
3.3 to 3.5 have the highest quality rating. 

```{r echo=FALSE}
ggplot(data=wine, aes(x=citric.acid, y=pH))+
  geom_point()

```


I find it interesting that cirtic acid and pH do not correlate as strongly as I 
had expected. After doing some reaserach, I found that "the most prevalent acids 
found in wine are tartaric acid, malic acid, and citric acid" (winefolly.com), 
and that these acids combined tend to influence the pH. Instead, volatile or 
fixed acidity may be better indicators of pH. 

```{r echo=FALSE}
ggplot(data=wine, aes(x=fixed.acidity, y=volatile.acidity))+
  geom_point(alpha=0.1)
```

```{r echo=FALSE}
ggplot(data=wine, aes(x=volatile.acidity, y=quality))+
  geom_point(alpha=0.1)
```

```{r echo=FALSE}
ggplot(data=wine, aes(x=fixed.acidity, y=quality))+
  geom_point(alpha=0.1)
```

First, I examined the relationship between fixed and volatile acidity to see if 
there was any correlation between the two variables. As discovered in the 
correlation matrix, the two variables have a correlation coefficient of 
-0.02269729, demonstrating no real evidence of a linear realtionship. The 
scatterplot of the two variables emphasizes the fact that there is no linear 
relationship as the data is clustered in the lower left portion of the graph. 

I then looked at the relationship of each variable individually with quality, 
and was unable to identify a clear realtionship with quality for either. 

```{r echo=FALSE}
ggplot(data=wine, 
       aes(x=fixed.acidity, y=pH))+
  geom_point(alpha=0.25)
```

Although the measures of acidity did not show a relationship with quality, a 
relationship is evident between fixed acidity and pH. As fixed acidity 
increases, the pH decreases, which makes sense, as a greater acidic level would 
indicate a lower value on the pH scale. 

```{r echo=FALSE}
ggplot(data=wine, 
       aes(x=residual.sugar, y=alcohol))+
  geom_point(alpha=0.1)
```

After adjusting for overplotting, the moderate negative relationship between 
alcohol and residual sugar and wine appears. It is more common for lower alcohol 
percentage wines to have higher sugar contents. 

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Alcohol and residual sugar have a moderate negative linear relationship. 

Alcohol content has a stronger correlation coefficient with quality than any 
other variable, with a pearson correlation coefficient value of 0.435574715. 
Moving into the multivariate analysis section, I will explore the effects of 
alcohol on quality in conjunction with other variables of interest. 

The variables describing the acidity of wine did not display much influence on 
the quality of wine, however, I did find that fixed acidity has the greatest 
impact on the pH of the wine. 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Density is highly correlated with both alcohol content and residual sugar. As 
residual sugar increases, the density of wine also increases. As the alcohol 
content increases, the density of the wine decreases. 

Also, I discovered that total sulfur dioxide and density have a moderate 
positive linear relationship. This was not necessarily something I was searching 
for in the dataset, but it is interesting to note as I explore density more 
thouroughly. 

### What was the strongest relationship you found?

The strongest correlation is between density and residual sugar, with a pearson 
correlation coefficient of 0.83896645, indicating that as the amount of residual 
sugar in the wine increases, the density increases relatively. 

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
ggplot(data=wine, 
       aes(x=pH, y=quality))+
  geom_point(aes(color=fixed.acidity), alpha=0.25)+
  scale_colour_gradientn(colours=rainbow(4))
```

In this plot, pH and fixed acidity maintain their negative relationship. The 
greater the fixed acidity, the lower the pH. However, I was still unable to 
identify a relationship between acidity and quality rating as I had initially 
expected. 

```{r echo=FALSE}
ggplot(data=wine, 
       aes(x=alcohol, y=density, colour=residual.sugar))+
  geom_point(alpha=.5)+
  scale_colour_gradientn(colours=rainbow(4))
  
```

Alcohol and residual sugar influence the density of wine, as discovered in the 
bivariate plots section. Alcohol and density have a negative linear 
relationship, while residual sugar and density have a positive linear 
relationship. The plot exhibits a decreasing trend, indicating the negative 
relationship between alcohol and density, and the higher sugar points fall 
higher on the plot, indicating the positve relationship between residual sugar 
and density. 


```{r echo=FALSE}
ggplot(data=wine, 
       aes(x=alcohol, y=density, colour=total.sulfur.dioxide))+
  geom_point(alpha=.5)+
  scale_color_gradient(low = "yellow", high = "black")
  
```

This plot follows a similar structure as the previous plot. Alcohol and total 
sulfur dioxide have a moderate negative linear relationship, while density and 
total sulfur dioxide have a moderate positive linear relationship. 

```{r echo=FALSE}
ggplot(data=wine, 
       aes(x=volatile.acidity, y=quality, colour=fixed.acidity))+
  geom_point()+
  scale_colour_gradientn(colours=rainbow(4))
```

I wanted to re-examine the relationship between fixed acidity, volatile acidity 
and quality of wine by observing all three variables on the same plot. However, 
the output simply re-enforces my previous conclusions of no significant 
relationship. 


```{r}
ggplot(data=wine, 
       aes(x=total.sulfur.dioxide, y=density, colour=free.sulfur.dioxide))+
  geom_point(alpha=0.2)+
  scale_color_gradient(low = "yellow", high = "black")
```

Density and total sulfur dioxide display a moderate positive relationship. As 
the total sulfur dioxie in wine increases, the density of the wine increases as 
well. This would also indicate that wines with higher total sulfur dioxide 
likely have a greater amount of residual sugar, and lower alochol percentages. 
Although, these relationships are simply a result of the relationship of each 
variable to density. I do not believe total sulfur dioxide would have a direct 
effect on alcohol or residual sugar content. 

As we would expect, this plot does display the positive relationship between 
total sulfur dioxide and free sulfur dioxide. Wines with higher values of total 
sulfur dioxide tend to have greater amounts of free sulfur dioxide. 

```{r echo=FALSE}
ggplot(data=wine, aes(x=alcohol, y=quality, colour=density))+
  geom_point()+
  scale_color_gradient(low = "yellow", high = "black")
 
```

I took a look at the plot between alcohol, quality and density as I recognized 
relationships between alcohol and quality, and alcohol and density. I wanted to 
explore whether I could view the interactions between the variables in a single 
plot. Although density and quality do not have a clear realtionship, the 
tendency of lower alcohol wines to have a greater density is apparent, with the 
darker colored points falling on the lower end of the x-axis. 

```{r}
ggplot(data=wine, 
       aes(x=alcohol, y=density, color=residual.sugar))+
  facet_wrap(~wine$quality)+
  geom_point(alpha=0.25)+
  scale_colour_gradientn(colours=rainbow(4))
```

The plots of alcohol, density and residual sugar are similar in nature across 
quality ratings. There is a significant point in the 6 rating plot, in which 
residual sugar is very high, which gives the wine a high density. Otherwise, 
across the ratings, as residual sugar is lower, the density is lower. Also, 
wines with higher amounts of residual sugar tend to have lower alcohol 
percentages. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The main relationships I explored were the ones between residual sugar, alcohol, 
and quality. Although pH was a feature of interest, it did not necessarily show 
a relationship with any of the other features of interest from a multivariate 
standpoint. 

The relationship between density, alcohol and residual sugar was very apparent 
when viewed in a single plot. As previously stated, density and alcohol 
percentage have a negative relationship, while density and residual sugar have a 
positive relationship. 

### Were there any interesting or surprising interactions between features?

I was very surprised that the acidity variables did not have a large influence 
on quality ratings. When researching wine tasting, I found that "Acidity gives 
wine its tart and sour taste" (Winefolly). Therefore, I assumed that measures of 
acidity (pH, citric acid, volatile acidity and fixed acidity) would have a clear 
effect on the wine quality. However, I was unsuccessful in identifying a 
relationship. 

I did discover a relationship between total sulfur dioxide and density. The two
variables have a moderate positive linear relationship. Although this was not
necessarily something I was searching for, it is interesting to note. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, Plot_One}
p1 <- ggplot(data=wine, aes(x=alcohol, y=quality))+
  geom_point(stat='summary', fun.y=mean)+
  geom_smooth(color='blue')+
  geom_smooth(method='lm', color='red')+
  ggtitle('Average Wine Quality by Alcohol Content')

p2 <- ggplot(data=subset(wine, !is.na(alcohol_bucket)), 
       aes(x=alcohol_bucket, y=quality))+
  geom_point(stat='summary', fun.y=mean)+
  ggtitle('Average Quality by Alcohol Content Range')

grid.arrange(p1, p2) 
  
```

### Description One

As discovered in the bivariate plots section, alcohol was the characteristic of 
wine with the greatest influence on quality. I wanted to develop a model that 
best depicted this relationship. Looking at the original scatterplot of alcohol 
content v. quality, it was very difficult to examine the relationship between 
the two variables. After adjusting for the noise in the bivariate section by 
cutting the variable into whole number percentages (e.g. (9,10]), I wanted to 
see how well this adjustment fit the original data. To do so, I used the 
original graph of average quality rating v. alcohol percentage, and added two 
different smoothers. The blue smoother was created using the default settings 
of the geom_smooth() function (method = 'gam' and formula 'y ~ s(x, bs = "cs")). 
The red smoother utilized the linear method. 

The red smoother exhibits an increasing trend, while the blue smoother exhibits 
an increasing trend just after 9% alcohol content. The blue smoother has the 
same initial dip as found in the plot of average quality rating by alcohol 
content range. If interested in prediciting the quality of wine solely from 
alcohol content, this may be the best model for said purpose. Another aspect of 
the plot that is made more apparent with the addition of the smoothers is that 
there is that the greatest variance in average quality occurs between 10 and 
12.5%. Further investigation would be necessary in determining the source of 
this higher variance. 

### Plot Two
```{r echo=FALSE, Plot_Two}
p3 <- ggplot(data=wine, 
       aes(x=alcohol, y=density, colour=residual.sugar))+
  geom_point(alpha=.5)+
  scale_colour_gradientn(colours=rainbow(4))+
  xlab('Alcohol % by Volume')+
  ylab('Density (g/cm^3)')+
  labs(col="Residual Sugar\n")
  ggtitle('Wine Density by Alcohol % and Residual Sugar')

p4 <- ggplot(data=wine, 
       aes(x=residual.sugar, y=density, colour=alcohol))+
  geom_point(alpha=.5)+
  scale_colour_gradientn(colours=rainbow(4))+
  xlab('Residual Sugar(g/dm^3)')+
  ylab('Density (g/cm^3)')+
  labs(col="Alcohol %\n")
  ggtitle('Wine Density by Residual Sugar and Alcohol %')

grid.arrange(p3, p4, ncol=2)
```

### Description Two

In the bivariate section, I initially examined the plot between density, alcohol 
percentage and residual sugar, with the points colored in accordance to the 
amount of residual sugar present in the wine. I wanted to examine the graph with 
residual sugar plotted against density, colored by alcohol percentage. Both 
graphs portray essentially the same concepts, yet it is useful to examine the 
information from both plots. 

We can see that alcohol percentage and residual sugar both have a linear 
correlation with the density of white wine. 

### Plot Three
```{r echo=FALSE, message=FALSE, Plot_Three}
wine_sub <- subset(wine, !is.na(wine$alcohol_bucket))

qplot(data=wine_sub, 
      x=pH, bins=10,
      ylab= 'Count')+
  facet_wrap(~quality)+
  ggtitle("pH Distribution by Quality Rating")
  
```

### Description Three

For my final plot, I wanted to re-examine the relationship between pH and 
quality. The plot provided earlier between the cut variable ph_buckets and 
quality displayed an ideal pH within the range of (3.3, 3.5]. However, the 
dataset description explains that the dataset is not balanced, and contains 
more mid-range quality wines, and fewer on the extreme ends of the spectrum 
(very high or low quality). This makes sense, as the average quality rating for 
a wine in the pH range of (3.3, 3.5] was just under 6.1. Therefore, I wanted to 
take a look at the pH distribution in terms of each quality rating, to determine 
whether or not this ideal range was consistent across quality ratings, or 
whether it simily stemmed from the mid-range wines. 

While these distributions do show that the mid-range wines have the greatest 
influence on the values, we can see that as the quality rating increases above 
the mid-rating, the pH distribution is less dispersed, and falls closer to the 
(3.3, 3.5] range. For instance, the distribution of the wines rated 7 ranges 
from about 2.75 to 3.8; for wines with a rating of 8 the distribution ranges 
from about 2.8 to 3.6; and the wines with a rating of 9 have the least spread in 
terms of pH, in which these fall approximately between 3.1 and 3.5. 

------

# Reflection

After my initial research on wine tasting and wine attributes, I was surprised 
by the lack of correlation I discovered in the dataset. There are other 
variables I would have liked to see in the dataset such as price of the wine, 
to test whether or not higher priced wines are typically rated at higher 
qualities. Additionally, it would be interesting to see how the wine experts 
percieved the wine. For instance, how accurately they can identify an acidic 
wine. 

Additionally, there are many questions that I have that can not necessarily be 
answered solely by this dataset. One pattern I am curious about is the heavy 
right skew throughout the dataset. I wonder why so many variables had high-value 
outliers. 

After examining the dataset, it can be concluded that alcohol has the greatest 
impact on the quality of white wine; the higher the alcohol content, the higher 
the quality rating, on average. 

In the future I would like to explore the red wine dataset to determine which 
variables are most significant in determining the quality of red wine, and 
whether or not the variables in the dataset show similar patterns to the white 
wine dataset. 


------

Resources: 

Understanding Acidity in Wine. (2015, December 09). Retrieved from https://winefolly.com/review/understanding-acidity-in-wine/ 

Wine Chemistry, Science Direct. (n.d.). Retrieved from https://www.sciencedirect.com/topics/agricultural-and-biological-sciences/
wine-chemistry 